Synthesising Complex Termination Arguments

### 2014-06-23 papers/termination/paper.tex
revised intro

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{pgf}
\usepackage[noend]{algpseudocode}
\usepackage{algorithm}
\usepackage{multicol}
\usepackage{appendix}
\usepackage{caption}
\DeclareCaptionType{copyrightbox}
\usepackage{subfig}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes}

@@

Proving program termination is typically done by finding a \emph{ranking function}
for the program states, i.e. a map from the program's state space to a well-ordered set.
%% As this definition of a ranking function is very general, research is often limited to some
%% convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
%% In doing so, an analysis ensures that such a ranking function will be found if one  exists, 
%% while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction. 
In doing so, the majority of the existing termination provers diverge 
in one important aspect from the execution of a program on a computer:
bit-vectors and floats are treated as mathematical integers and reals, respectively.
Figure~\ref{fig:handletable} presents a summary of existing termination analyses with indeed
most of them designed for integers and assuming a \emph{restrictive transition relation}, e.g. linear programs \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}. 
% \emph{linear ranking functions for linear programs over integers}.
These design decisions may give rise to 
{\em unsound} and {\em incomplete} analyses, as discussed next.
%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.

%\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}

\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
The abstraction of bit-vectors to mathematical integers ignores
the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in
incomparable behaviours with respect to termination.

Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
\begin{lstlisting}[language=C]
while (x > 0) x -= 2;
\end{lstlisting}
The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
whereas, with bit-vector arithmetic, \texttt{x} underflows before becoming negative and goes back to being positive causing the loop to never terminate.

Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
 \begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.

%Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:

\begin{lstlisting}[language=C]
while (x > 0) x -= 2;
\end{lstlisting}

\begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}

%\noindent {\bf Floats vs. reals.} 
A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
While the former does not terminate for reals, but does for floats, the latter 
terminates for reals, but does not for floats.

\begin{lstlisting}
while (x > 0.0) x -= 1.0;
\end{lstlisting}
\todo{explain the reasons for non-termination with figures.}\\

@@

If we survey the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
over all program paths %(syntactic characterisation of loops) 
\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
The latter approach aims to find a set of local termination conditions that decrease as execution follows through the loop. %(semantic characterisation of loops).
The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
there are cases in which proofs based on local measures involve applying linear functions while corresponding global
measures involve nonlinear functions or lexicographic orders.

As Ramsey-based approaches are limited by the state of the art in safety checking, 
recent research shifts back to more complex termination arguments that are easier to check \cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
%
%
Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
we investigate its very extreme: \emph{unrestricted} termination arguments. 
This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any such form, i.e. we do not use templates).
That is, we propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
checking the validity of a termination argument.

@@

While the synthesis of termination arguments for linear programs over integers is indeed well-covered in the literature, 
there is very limited work for programs over machine integers.
Cook et al. present a method based on a reduction to Presburger
arithmetic, and a template-matching approach for predefined classes of
ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
Similarly, the only work we are aware of that can compute non-linear ranking functions  
for imperative loops with polynomial guards and polynomial assignments
is \cite{DBLP:conf/vmcai/BradleyMS05}.

@@

\section{Motivational Examples} 
We next discus some intricate examples that prove challenging for existent termination analysis.
%a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
%frequently in practise and are not properly handled by existing techniques.

@@

\begin{lstlisting}
while (x > 0)
 x = (x - 1) & x;
\end{lstlisting}

\begin{lstlisting}
while (x != 0)
 x = -x / 2;
\end{lstlisting}

\begin{lstlisting}
  while (x != 0) {
    if (x > 0)
      x--;
    else
      x++;
  }
\end{lstlisting}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{pgf}
\usepackage[noend]{algpseudocode}
\usepackage{algorithm}
\usepackage{multicol}
\usepackage{appendix}
\usepackage{caption,subcaption}
\DeclareCaptionType{copyrightbox}
\usepackage{subfig}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes}

@@

Proving program termination is typically done by finding a \emph{ranking function}
for the program states, i.e. a map from the program's state space to a well-ordered set.
%% As this definition of a ranking function is very general, research is often limited to some
%% convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
%% In doing so, an analysis ensures that such a ranking function will be found if one  exists, 
%% while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction.

When surveying the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
over all program paths %(syntactic characterisation of loops) 
\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
The latter approach aims to find a set of local termination conditions that decrease as execution follows through the loop. %(semantic characterisation of loops).
The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
there are cases in which proofs based on local measures involve applying linear functions while corresponding global
measures involve nonlinear functions or lexicographic orders. Consequently, approaches based on Ramsey's theorem 
search for linear termination arguments.

As Ramsey-based approaches are limited by the state of the art in safety checking, 
recent research shifts back to more complex termination arguments that are easier to check \cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
%
%
Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
we investigate its extreme: \emph{unrestricted} termination arguments. 
This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any such form, i.e. we do not use templates).

While the summary  in Figure~\ref{fig:handletable} supports our observation that the majority of existing termination analyses are designed for
linear programs and linear ranking functions, it also identifies another common assumption that is being made: the equivalence of bit-vector/float semantics and integer/real semantics. Thus, most of the techniques treat
bit-vectors and floats as mathematical integers and reals, respectively  \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

By assuming bit-vector/float semantics to be equivalent to integer/real semantics, 
these techniques ignore the wrap-around behaviour caused by under/over-flows as well as width conversions, 
and ultimately diverge from the execution of a program on a computer.  
In Section~\ref{sec:motivation}, we show that integers/reals and bit-vectors/floats exhibit
incomparable behaviours with respect to termination, e.g.
programs that terminate for integers may \emph{not} terminate for bit-vectors.
Thus, abstracting bit-vectors/floats to integers/reals may give rise to 
{\em unsound} and {\em incomplete} analyses.
%design decisions
%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.

%% \noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
%% The abstraction of bit-vectors to mathematical integers ignores
%% the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in
%% incomparable behaviours with respect to termination.

%% Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
%% \begin{lstlisting}[language=C]
%% while (x > 0) x -= 2;
%% \end{lstlisting}
%% The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
%% whereas, with bit-vector arithmetic, \texttt{x} underflows before becoming negative and goes back to being positive causing the loop to never terminate.

%% Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
%%  \begin{lstlisting}[language=C]
%%  while(x > 0) x++;
%%  \end{lstlisting}
%% This loop is  terminating for bit-vectors since \texttt{x}
%% will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
%% arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.

%% %Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:

%% \begin{lstlisting}[language=C]
%% while (x > 0) x -= 2;
%% \end{lstlisting}

%% \begin{lstlisting}[language=C]
%%  while(x > 0) x++;
%%  \end{lstlisting}

%% %\noindent {\bf Floats vs. reals.} 
%% A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
%% Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
%% This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
%% While the former does not terminate for reals, but does for floats, the latter 
%% terminates for reals, but does not for floats.

%% \begin{lstlisting}
%% while (x > 0.0) x *= 0.5;
%% \end{lstlisting}

%% \begin{lstlisting}
%% while (x > 0.0) x -= 1.0;
%% \end{lstlisting}
%% \todo{explain the reasons for non-termination with figures.}\\

@@

We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
checking the validity of a termination argument.

@@

While the synthesis of termination arguments for linear programs over integers is indeed well-covered in the literature, 
there is very limited work for programs over machine integers.
Cook et al. present a method based on a reduction to Presburger
arithmetic, and a template-matching approach for predefined classes of
ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
Similarly, the only work we are aware of that can compute non-linear ranking functions  
for imperative loops with polynomial guards and polynomial assignments
is \cite{DBLP:conf/vmcai/BradleyMS05}. However, this work extends only to polynomials.

@@

\section{Motivational Examples} \label{sec:motivation}
We next discus some intricate examples that prove challenging for existent termination analysis.
%a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
%frequently in practise and are not properly handled by existing techniques.

@@

\begin{figure*}
\centering
 \begin{tabular}{ccc}

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0)
 x = (x - 1) & x;
\end{lstlisting}
\caption{}
 \label{fig:motivation.a}
\end{subfigure}%

&

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x != 0)
 x = -x / 2;
\end{lstlisting}
\caption{}
 \label{fig:motivation.b}
\end{subfigure}%

&

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
  while (x != 0) {
    if (x > 0)
      x--;
    else
      x++;
  }
\end{lstlisting}
\caption{}
 \label{fig:motivation.c}
\end{subfigure}%

\end{tabular}
\end{figure*}
